Adapting the MVDR to a GPU for active sonar systems of different sizes.

### 2013-01-22 ASA2013/asa2013.tex
la2pdf commit - 2013:01:22	16:54:53	(Tue22.Jan)

Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome. The MVDR is a technique that displays a great potential for improving image quality over simple conventional methods, but is far more computationally complex. This is because a spatial covariance matrix must be estimated (of O($M$) and inverted for each image pixel.  For systems with less than 20-30 channels, the estimation step dominates,   %Unlike other studies, w While most studies tend to focus on reducing the dimensionality of the convariance matrix to

@@

The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
\\
<< MOTIVATION >>
\\
Comparable can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a,}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
\\
This article is outlined as follows: In section \ref{methods} we will present the theoretical background on MVDR operated in time and frequency domain, as well as relating this to the LCA beamformer.  As we move along we aim to mention complexity issues and suggest means to overcome them, but implementational aspect will first be presented in \ref{implementation}.

@@

Suppose that a backscattered wavefield is recorded by an $M$ element uniform linear array, that a suitable set of delays are applied to the channels for the pixel of interest, and let the delayed output from the $m$'th channel be given as $x_m$. By definition a beamformer's output $z[n]$ at time instant $n$ is a weighted sum all the delayed data outputs:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
\end{align}
where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. Various weighting patterns exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two\todo{don't quite like this one}~\cite{Harris1978}.

@@

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=0.7\linewidth]{gfx/implementation.svg}%
\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). }\label{mvdr_beamforming}
\end{figure}
Adaptive beamformers target this limitation by computing a set of weights for each pixel that to better fit the dynamic nature of the incoming wavefield. The MVDR beamformer is one such method. It computes a set of weights that minimizes the noise and signal power accumulated by the array, but with a constraint of unity gain in the desired direction. When the data is delayed this is formulated as follows:
\begin{gather}
\argmin{\w}\ E\Big\{\big|\w\H[n]\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w\H[n]\1 = 1. \label{mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
This is a convex optimization problem with the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array. Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$. In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}. Combined, these techniques will also ensure a numerically well conditioned $\eR$.

>>>>

Through a case study of on such technique, the Minimum Variance Distortionless Response (MVDR) beamformer, we identify these challenges and explain how they were overcome. The MVDR is a technique that displays a great potential for improving image quality over simple conventional methods, but is far more computationally complex. This is because a spatial covariance matrix must be estimated and inverted for each image pixel.  For systems with less than 20-30 channels, the estimation step dominates,   %Unlike other studies, w While most studies tend to focus on reducing the dimensionality of the convariance matrix to

@@

The end design is more than 2 orders of magnitude faster than a C implementation we started off with, it took considerable effort to get there.
\\
<< MOTIVATION >>
\\
Comparable can be found e.g. in ultrasound imaging, where Chen et. al have investigated porting a time domain MVDR method to a GPU \cite{Chen2009,Chen2011,Chen2011a,}. Unlike their work, however, we do not impose the restriction of the input data to be real, nor do we approach larger systems by enforcing Toeplitz structure of the covariance matrix. 
\\
This article is outlined as follows: In section \ref{methods} we will present the background on the MVDR method operated in time and frequency domain, and on the LCA beamformer. For each method we start with the mathematical formulation, then evaluate complexity issues, and finally give our thoughts on some implementational aspects. The results are collected in \ref{implementation}. Finally, we conclude in \ref{discssion}.

@@

Suppose that a backscattered wavefield is recorded by an $M$ element uniform linear array, that a suitable set of delays are applied to the channels for the pixel of interest, and let the delayed output from the $m$'th channel be given as $x_m$. By definition a beamformer's output $z[n]$ at time instant $n$ is a weighted sum all the delayed data outputs:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
\end{align}
where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. Various weighting patterns exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two\todo{don't quite like this one}~\cite{Harris1978}. Adaptive beamformers target this limitation by computing a set of weights for each pixel that to better fit the dynamic nature of the incoming wavefield.

@@

\begin{figure}[t]\centering%
\graphicsAI[drawing,width=0.7\linewidth]{gfx/implementation.svg}%
\caption{MVDR beamforming. For one of the total number of pixels in range and azimuth, $N_y$ and $N_x$,\newline
1. an $L\times{}L$ sample covariance matrix $\eR$ is computed, \newline
2. the term $\eR^{-1}\1$ is found using a linear equation solver,\newline
3. and the beamformer output is computed from $z$ from (\ref{finalZ}), where $\w$ is found by substituting $\eR^{-1}\1$ into (\ref{weights}). }\label{mvdr_beamforming}
\end{figure}
The MVDR beamformer is one such method. It computes a set of weights that minimizes the noise and signal power accumulated by the array, but with a constraint of unity gain in the desired direction. When the data is delayed this is formulated as follows:
\begin{gather}
\argmin{\w}\ E\Big\{\big|\w\H[n]\x[n]\big|^2\Big\} \qquad\text{subject to}\qquad \w\H[n]\1 = 1. \label{mvdr_def}
% \vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
This is a convex optimization problem with the solution:
\begin{gather}
\vec w[n] = \frac{\Ri[n]\1}{\1\T\Ri[n]\1},\label{weights}
\end{gather}
where $\1$ is a row vector of ones that represents broadside steering, and $\R=E\{\x[n]\x\H[n]\}$ is the spatial covariance matrix for the full array. Since $\R$ is unknown, we estimate it by by computing a sample covariance matrix $\eR$. In this computation we perform some degree of \emph{spatial averaging} to avoid signal cancellation, \emph{temporal averaging} to maintain true speckle statistics, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Synnevag2009a}. Combined, these techniques will also ensure a numerically well conditioned $\eR$.
